Accuracy in Parameter Estimation and Simulation Approaches for Sample Size Planning with Multiple Stimuli

### 2023-12-28 manuscript_draft.tex
so many updates

\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same}
\hypersetup{
  pdftitle={Accuracy in Parameter Estimation and Simulation Approaches for Sample Size Planning with Multiple Stimuli},
  pdfauthor={Erin M. Buchanan1 \& Other Folks as Per Order on Doc2},
  pdflang={en-EN},
  pdfkeywords={accuracy in parameter estimation, power, sampling, simulation, hypothesis testing},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

\title{Accuracy in Parameter Estimation and Simulation Approaches for Sample Size Planning with Multiple Stimuli}
\author{Erin M. Buchanan\textsuperscript{1} \& Other Folks as Per Order on Doc\textsuperscript{2}}
\date{}

@@

The authors made the following contributions. Erin M. Buchanan: Conceptualization, Writing - Original Draft Preparation, Writing - Review \& Editing, Resources, Validation, Visualization, Project Administration, Formal Analysis; Other Folks as Per Order on Doc: Writing - Review \& Editing, Data Curation, Resources.

Correspondence concerning this article should be addressed to Erin M. Buchanan, 326 Market St., Harrisburg, PA, 17101. E-mail: \href{mailto:ebuchanan@harrisburgu.edu}{\nolinkurl{ebuchanan@harrisburgu.edu}}

@@

\affiliation{\vspace{0.5cm}\textsuperscript{1} Harrisburg University of Science and Technology\\\textsuperscript{2} Other Instituions}

@@

Statistical power and power analyses are arguably one of the most important components in planning a research study (Cohen, 1990). However, if reviews of transparency and openness in research publications are any clue, the social sciences have failed to fully implement power analyses as part of their common efforts (Hardwicke et al., 2020, 2022). The replication ``crisis'' and credibility revolution have shown that published studies in psychology are underpowered (Open Science Collaboration, 2015; Vazire, 2018). Pre-registration of a study involves outlining the study and hypotheses before data collection begins (Chambers et al., 2014; Nosek \& Lakens, 2014; Stewart et al., 2020), and details of a power analyses or other limitations on resources are often used to provide justification for the pre-registered sample quota. Given the combined issues of publish-or-perish and that most non-significant results do not result in published manuscripts, one may expect that power analysis would be especially critical for early career researchers (Rosenthal, 1979; Simmons et al., 2011). At best, an underpowered study provides limited insight (Halpern, 2002), and it can be difficult to know if a poorly implemented power analysis is better than no power analysis.

A recent review of power analyses found - across psychology journal articles - researchers did not provide enough information to understand their power analyses and often chose effect sizes that were unjustified (Beribisky, 2019). One solution to this power analysis problem is the plethora of tools made available for researchers to make power computations accessible to non-statisticians with the caveat that a solid education in power is necessary to use these tools. G*Power is one of the most popular free power software options (Erdfelder et al., 1996; Faul et al., 2007) that provides a simple point and click graphical user interface for power calculations (however, see Brysbaert, 2019). Web-based tools have also sprung up for overall and statistical test specific sample size planning including \url{https://powerandsamplesize.com} and \url{https://designingexperiments.com} (Anderson et al., 2017). \emph{R}-coding based packages, such as \emph{pwr} (Champely et al., 2017), \emph{faux} (DeBruine, 2021), and \emph{SimDesign} (Chalmers \& Adkins, 2020), can be used to examine power and sample size planning usually with simulation. Researchers must be careful using any toolkit, as errors can occur with the over-reliance on software (e.g., it should not be a substitute for critical thinking, Nuijten et al., 2016). When computing sample size estimates, it is important to remember that these values are estimations, not exact calculations guaranteed to produce a specific result (Batterham \& Atkinson, 2005). For example, if any parameter estimated by the researcher was not found in the study (i.e., a smaller effect size than used for the power analysis), then the sample size estimate tied to that specific level of power may be incorrect.

Changes in publication practices and research design have also created a new wrinkle in providing a sample size plan for a research study. While statistics courses often suggest that a specific research design leads to a specific statistical test, meta-science work has shown that given the same data and hypothesis, researchers can come up with multiple ways to analyze the data Coretta et al. (2023). Therefore, a single power analysis only supports the specific analysis in which the researcher expects to test. Analyses may evolve during the research project or be subject to secondary analysis; thus, power and sample size estimation based on one analysis is potentially less useful than previously imagined. Further, research projects often have multiple testable hypotheses, but it is unclear which hypothesis or test should be used to estimate sample size with a power analysis. Last, research publications may not even have a specific, testable hypothesis, as some publications are intended to curate a large dataset for future reuse (i.e., stimuli database creation, Buchanan et al., 2019).

In light of these analytical (or lack thereof) concerns, we propose a new method that combines accuracy in parameter estimation Maxwell et al. (2008) and bootstrapped simulation on pilot data (Rousselet et al., 2022). This method accounts for a potential lack of hypothesis test (or simply not a good way to estimate an effect size of interest), and/or an exploratory design with an unknown set of potential hypotheses and analytical choices. Additionally, we consider the nature of cognitive research designs that use multiple items to measure the phenomena of interest. For example, semantic priming is measured with multiple paired stimuli (Meyer \& Schvaneveldt, 1971), which traditionally was analyzed by creating person or item-level averages for an ANOVA (Brysbaert \& Stevens, 2018). However, the use of multilevel models with random effects for the stimuli used in a study have shown that we should be careful to assume that all items of a research study have the same ``effect'', as there is often variability in their impact on the outcomes of the study.

@@

AIPE shifts the focus away from finding a significant p-value to finding a parameter that is ``accurately measured''. For example, researchers may wish to detect a specific correlation in a study, \emph{r} = .35. They could then use AIPE to estimate the sample size needed to find a ``sufficiently narrow'' confidence interval around that correlation. Sufficiently narrow is often defined by the researcher using a minimum parameter size of interest and confidence intervals. Therefore, they could decide that their 95\% confidence interval should be approximately between .20 and .50, and sufficiently narrow could be defined as a width of .30 or .15 on each side. While confidence intervals are related to null hypothesis significance testing (i.e., 95\% confidence intervals that do not include zero would indicate a significant difference from zero at \(\alpha\) \textless{} .05), AIPE procedures suggest how we can define a sample size with a given width of confidence interval, regardless of whether it includes zero.

@@

Bootstrapping involves using data obtained to simulate similar datasets by drawing from the original data with replacement (Efron, 2000; Rousselet et al., 2022). Bootstrapping is a form of data simulation that allows one to calculate parameter estimates, confidence intervals, and more to simulate the potential population distribution, shape, and bias. Simulation is often paired with making up data for testing analyses and hypotheses based on proposed effect size or suggested population means. Generally, we would suggest starting with pilot data of a smaller sample size to understand the variability in potential items used to represent your phenomenon, especially if they are to be used in a larger study. However, given some background knowledge about the potential items, one could simulate example pilot data to use in. similar manner in our suggested procedure. Pilot or simulated data would be used to estimate the variability within items and select a ``sufficiently narrow'' window for item's confidence interval for AIPE (i.e., by selecting a specific standard error criterion, given the formula for confidence intervals). Bootstrapping would then be used to determine how many participants may be necessary to achieve a dataset wherein many items meet the required confidence interval.

@@

Researchers could then use sequential testing to estimate their parameter of interest after each participant's data was collected to determine whether they have achieved their expected width of the confidence interval around that parameter. One would set a minimum sample size (i.e., based on known data collection ability) and use the confidence interval width as a stopping rule (i.e., stop data collection when the CI is narrow, as defined above). Next, researchers would use the estimated sample size associated with the simulation results of many items obtaining the stopping rule as a maximum sample size (i.e., we expect 90\% of items to meet our stopping rule with 100 participants based on simulation). By defining each of these components, researchers could ensure a feasible minimum sample size, a way to stop data collection when goals have been met, and a maximum sample size rule to ensure an actual end to data collection. Therefore, we proposed that we should be able to leverage the ideas behind AIPE, paired with simulation and bootstrapping, to estimate the minimum and maximum proposed sample sizes and stopping rules for repeated measures studies with expected variability in parameter estimates for items.

\hypertarget{proposed-method-to-calculate-sample-size}{%
\subsection{Proposed Method to Calculate Sample Size}\label{proposed-method-to-calculate-sample-size}}

Using these ideas, we suggest the following procedure to determine a sample size for each item:

\begin{enumerate}
\def\labelenumi{\arabic{enumi})}
\item
  Use pilot data that closely resembles data you intend to collect. This dataset should contain items that are identical or similar to those that will be collected in the study. In this procedure, it is important to ensure that the data is representative of a larger population of sampled items that you intend to assess. Generally, pilot data sample sizes will be small, as the goal would be to determine how many items would be necessary to reach a ``stable'' standard error.
\item
  Calculate the standard error (SE) in the subset of pilot data associated with each item to create a cutoff score that defines when items are considered ``accurately measured''. Next, the simulations below will explore what criterion should be used to determine the cutoff score from the pilot data.
\item
  Sample, with replacement, from your pilot data using sample sizes starting at 20 participants and increase in small units up to a value that you consider the maximum sample size. We will demonstrate example maximum sample sizes based on the data simulation below; however, a practical maximum sample size may be determined by time (e.g., one semester data collection) or resources (e.g., 200 participants worth of funding). Although 20 participants would likely yield imprecise estimates, we suggest this starting minimum for simulation purposes.
\item
  For each simulation, calculate the standard error for each item, and use these values to determine the percentage of items that meet the cutoff score determined in Step 2.
\item
  Find the minimum sample size so that 80\%, 85\%, 90\%, and 95\% of the items meet the cutoff score and can be considered accurately measured. We recommend these scores to ensure that most items are accurately measured, in a similar vein to the common power-criterion suggestions. Each researcher can determine which of these is their minimum or maximum sample size (e.g., individuals can choose to use 80\% as a minimum and 90\% as a maximum or use values from Step 3 based on resources).
\item
  Report these values, and designate a minimum sample size, the cutoff criterion, and the maximum sample size. Each researcher should also report if they plan to use an adaptive design, which would stop data collection after meeting the cutoff criterion for each item.
\end{enumerate}

These steps are summarized in Table \ref{tab:table-summary}. We will first demonstrate the ideas behind the steps using open data (Balota et al., 2007; Brysbaert et al., 2014). This example will reveal a few areas of needed exploration for the steps. Next, we portray simulations for the proposed procedure and find solutions to streamline and improve the sample size estimation procedure. Finally, we include additional resources for researchers to use to implement the estimation procedure.

@@

In this section, we provide two examples of the suggested procedure. The first example includes concreteness ratings from Brysbaert et al. (2014). Instructions given to participants denoted the difference between concrete (i.e., ``refers to something that exists in reality'') and abstract (i.e., ``something you cannot experience directly through your senses or actions'') terms. Participants were then asked to rate concreteness of terms using a 1 (\emph{abstract}) to 5 (\emph{concrete}) scale. This data represents a small scale dataset that could be used as pilot data for a study using concrete word ratings. The data is available at \url{https://osf.io/qpmf4/}.

The second dataset includes a large scale dataset with response latencies, the English Lexicon Project (Balota et al., 2007). The English Lexicon Project consists of lexical decision response latencies for English words. In a lexical decision task, participants simply select ``word'' for real words (e.g., \emph{dog}) and ``nonword'' for pseudowords (e.g., \emph{wug}). The trial level data is available here: \url{https://elexicon.wustl.edu/}. Critically, in each of these examples, the individual trial level data for each item is available to simulate and calculate standard errors on. Data that has been summarized could potentially be used, as long as the original standard deviations for each item were present. From the mean and standard deviation for each item, a simulated pilot dataset could be generated for estimating new sample sizes. All code to estimate sample sizes is provided on our OSF page, and this manuscript was created with a \emph{papaja} (Aust et al., 2022) formatted Rmarkdown document.

For this example, our researcher wants to determine the differences in response latencies for abstract and concrete words. They will select \emph{n} = 40 words from the rating data that are split evenly into abstract and concrete ends of the rating scale. In the experiment, each participant will rate the words for their concreteness, and then complete a lexical decision task with these words as the object of interest. Using both datasets, we can determine the sample size necessary to ensure adequately measured ratings and response latencies.

\emph{Step 1}. The concreteness ratings data includes 27250 concepts that were rated for their concreteness. We randomly selected \emph{n} = 20 abstract words (\(M_{Rating}\) \textless= 2) and \emph{n} = 20 concrete words (\(M_{Rating}\) \textgreater= 4). In the original study, not every participant rated every word, which created uneven sample sizes for each word. Further, participants were allowed to indicate they did not know a word, and this data was set to missing data. In our sample of 40 words, the average pilot sample size was 28.23 (\emph{SD} = 1.35), and we will use 28 as our pilot sample size for this example.

The ELP response latency data includes 27250 word-forms, 219 that are listed as non-words, and 27031 real words. We selected the same words as the concreteness subset selected above. The average pilot sample size for this random sample was 32.80 (\emph{SD} = 0.56), and \emph{n} = 33 will be our pilot size for this example.

@@

\emph{Step 3-5}. The pilot data was then simulated, with replacement, with samples from 20 to 300 increasing in units of 5, separately for concreteness and lexical decision times (Step 3). The standard error of each item was calculated for the bootstrapped samples (Step 4), and the percentage of items below each potential cutoff was gathered (Step 5). The smallest sample size with at least 80\%, 85\%, 90\%, and 95\% of items below the cutoff are reported in Table \ref{tab:table-example} (Step 5).

@@

\caption{\label{tab:table-example}Sample Size Estimates by Decile for Concreteness Example}

>>>>

\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same}
\hypersetup{
  pdftitle={Accuracy in Parameter Estimation and Simulation Approaches for Sample Size Planning with Multiple Stimuli},
  pdfauthor={Erin M. Buchanan1, Mahmoud M. Elsherif2, Jason Geller3, Chris L. Aberson4, Necdet Gurkan5, Ettore Ambrosini6,26, Tom Heyman7, Maria Montefinese8, Wolf Vanpaemel9, Krystian Barzykowski10, Carlota Batres11, Katharina Fellnhofer12, Guanxiong Huang13, Joseph McFall14,27, Gianni Ribeiro15, Jan P. Röer16, José L. Ulloa17, Timo B. Roettger18, K. D. Valentine19,28, Antonino Visalli20, Kathleen Schmidt21, Martin R. Vasilev22, Giada Viviani23, Jacob F. Miranda24, \& Savannah C. Lewis25},
  pdflang={en-EN},
  pdfkeywords={accuracy in parameter estimation, power, sampling, simulation, hypothesis testing},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

\title{Accuracy in Parameter Estimation and Simulation Approaches for Sample Size Planning with Multiple Stimuli}
\author{Erin M. Buchanan\textsuperscript{1}, Mahmoud M. Elsherif\textsuperscript{2}, Jason Geller\textsuperscript{3}, Chris L. Aberson\textsuperscript{4}, Necdet Gurkan\textsuperscript{5}, Ettore Ambrosini\textsuperscript{6,26}, Tom Heyman\textsuperscript{7}, Maria Montefinese\textsuperscript{8}, Wolf Vanpaemel\textsuperscript{9}, Krystian Barzykowski\textsuperscript{10}, Carlota Batres\textsuperscript{11}, Katharina Fellnhofer\textsuperscript{12}, Guanxiong Huang\textsuperscript{13}, Joseph McFall\textsuperscript{14,27}, Gianni Ribeiro\textsuperscript{15}, Jan P. Röer\textsuperscript{16}, José L. Ulloa\textsuperscript{17}, Timo B. Roettger\textsuperscript{18}, K. D. Valentine\textsuperscript{19,28}, Antonino Visalli\textsuperscript{20}, Kathleen Schmidt\textsuperscript{21}, Martin R. Vasilev\textsuperscript{22}, Giada Viviani\textsuperscript{23}, Jacob F. Miranda\textsuperscript{24}, \& Savannah C. Lewis\textsuperscript{25}}
\date{}

@@

Authorship order was determined by tier: 1) Lead author, 2) authors who wrote vignettes, 3) authors who contributed datasets, 4) authors who contributed to conceptualization/writing, and 5) project administration team. Within these tiers individuals were ordered by number of CRediT contributions and then alphabetically by last name. Data curation was defined as writing vignettes, and resources was defined by submitting datasets with their metadata. All other CRediT categories are their traditional interpretation.

The authors made the following contributions. Erin M. Buchanan: Conceptualization, Data curation, Formal analysis, Investigation, Methodology, Project administration, Resources, Software, Validation, Visualization, Writing - original draft, Writing - review \& editing; Mahmoud M. Elsherif: Data curation, Resources, Writing - original draft, Writing - review \& editing; Jason Geller: Data curation, Resources, Writing - original draft, Writing - review \& editing; Chris L. Aberson: Data curation, Writing - original draft, Writing - review \& editing; Necdet Gurkan: Data curation, Writing - review \& editing; Ettore Ambrosini: Resources, Writing - original draft, Writing - review \& editing; Tom Heyman: Resources, Writing - original draft, Writing - review \& editing; Maria Montefinese: Resources, Writing - original draft, Writing - review \& editing; Wolf Vanpaemel: Resources, Writing - original draft, Writing - review \& editing; Krystian Barzykowski: Data curation, Resources, Writing - original draft, Writing - review \& editing; Carlota Batres: Resources, Writing - review \& editing; Katharina Fellnhofer: Resources, Writing - original draft, Writing - review \& editing; Guanxiong Huang: Resources, Writing - original draft, Writing - review \& editing; Joseph McFall: Resources, Writing - review \& editing; Gianni Ribeiro: Resources, Writing - original draft, Writing - review \& editing; Jan P. Röer: Resources, Writing - original draft, Writing - review \& editing; José L. Ulloa: Resources, Writing - original draft, Writing - review \& editing; Timo B. Roettger: Visualization, Formal analysis, Writing - original draft, Writing - review \& editing; K. D. Valentine: Conceptualization, Writing - original draft, Writing - review \& editing; Antonino Visalli: Writing - original draft, Writing - review \& editing; Kathleen Schmidt: Writing - original draft, Writing - review \& editing; Martin R. Vasilev: Writing - original draft, Writing - review \& editing; Giada Viviani: Writing - original draft, Writing - review \& editing; Jacob F. Miranda: Project administration, Writing - original draft, Writing - review \& editing; Savannah C. Lewis: Project administration, Writing - original draft, Writing - review \& editing.

Correspondence concerning this article should be addressed to Erin M. Buchanan, Enter postal address here. E-mail: \href{mailto:ebuchanan@harrisburgu.edu}{\nolinkurl{ebuchanan@harrisburgu.edu}}

@@

\affiliation{\vspace{0.5cm}\textsuperscript{1} Harrisburg University of Science and Technology\\\textsuperscript{2} University of Leicester, Department of Vision Sciences\\\textsuperscript{3} Department of Psychology, Princeton University\\\textsuperscript{4} Illumin Analytics\\\textsuperscript{5} Stevens Institute of Technology\\\textsuperscript{6} Department of Neuroscience, University of Padova, Padova, Italy\\\textsuperscript{7} Methodology and Statistics Unit, Institute of Psychology, Leiden University\\\textsuperscript{8} Department of Developmental and Social Psychology, University of Padova, Italy\\\textsuperscript{9} University of Leuven\\\textsuperscript{10} Applied Memory Research Laboratory, Institute of Psychology, Jagiellonian University\\\textsuperscript{11} Franklin and Marshall College\\\textsuperscript{12} ETH Zürich\\\textsuperscript{13} Department of Media and Communication, City University of Hong Kong\\\textsuperscript{14} Department of Psychology, University of Rochester\\\textsuperscript{15} School of Psychology, The University of Queensland\\\textsuperscript{16} Department of Psychology and Psychotherapy, Witten/Herdecke University\\\textsuperscript{17} Programa de Investigación Asociativa (PIA) en Ciencias Cognitivas, Centro de Investigación en Ciencias Cognitivas (CICC), Facultad de Psicología, Universidad de Talca\\\textsuperscript{18} University of Oslo\\\textsuperscript{19} Massachusetts General Hospital\\\textsuperscript{20} IRCCS San Camillo Hospital (Venice Lido, Italy)\\\textsuperscript{21} Ashland University\\\textsuperscript{22} Bournemouth University\\\textsuperscript{23} University of Padova\\\textsuperscript{24} California State University East Bay\\\textsuperscript{25} University of Alabama, Tuscaloosa\\\textsuperscript{26} Padova Neuroscience Center, University of Padova, Padova, Italy\\\textsuperscript{27} Children's Institute, Inc\\\textsuperscript{28} Harvard Medical School}

@@

An inevitable decision in almost any empirical research is deciding on the sample size. Statistical power and power analyses are arguably some of the most important components in planning a research study and its corresponding sample size (Cohen, 1990). However, if reviews of transparency and openness in research publications are any clue, researchers in the social sciences commonly fail to implement proper power analyses as part of their research workflow (Hardwicke et al., 2020, 2022). The replication ``crisis'' and credibility revolution have shown that published studies in psychology are underpowered (Korbmacher et al., 2023; Open Science Collaboration, 2015; Vazire, 2018). Pre-registration of a study involves outlining the study and hypotheses before data collection begins (Chambers et al., 2014; Nosek \& Lakens, 2014; Stewart et al., 2020), and details of a power analyses or limitations on resources are often used to provide justification for the pre-registered sample quota (Pownall et al., 2023; van den Akker, Assen, et al., 2023; van den Akker, Bakker, et al., 2023). Given the combined issues of publish-or-perish and that most non-significant results do not result in published manuscripts, power analysis may be especially critical for early career researchers to increase the likelihood that they will identify significant effects if they exist (Rosenthal, 1979; Simmons et al., 2011). Justified sample sizes may allow for publication of non-significant, yet practically important effects, along with the smallest effect of interest movement (Anvari \& Lakens, 2021), potentially improving the credibility of published work. At best, an underpowered study provides limited insight (Halpern, 2002), and it can be difficult to know if a poorly implemented power analysis is better than no power analysis.

A recent review of power analyses found - across behavioral, cognitive, and social science journal articles - researchers did not provide enough information to understand their power analyses and often chose effect sizes that were unjustified (Beribisky, 2019). One solution to this power analysis problem is the plethora of tools made available for researchers to make power computations accessible to non-statisticians; however, a solid education in power is necessary to use these tools properly. G*Power is one of the most popular free power software options (Erdfelder et al., 1996; Faul et al., 2007) that provides a simple point and click graphical user interface for power calculations (however, see Brysbaert, 2019). Web-based tools have also sprung up for overall and statistical test specific sample size planning including \url{https://powerandsamplesize.com}, \url{https://jakewestfall.shinyapps.io/pangea/}, \url{https://pwrss.shinyapps.io/index/}, and \url{https://designingexperiments.com} (Anderson et al., 2017). \emph{R}-coding based packages, such as \emph{pwr} (Champely et al., 2017), \emph{faux} (DeBruine, 2021), \emph{simr} (Green \& MacLeod, 2016), \emph{mixedpower} (Kumle \& DejanDraschkow, 2020), and \emph{SimDesign} (Chalmers \& Adkins, 2020), can be used to examine power and plan sample sizes, usually with simulation. Researchers must be careful using any toolkit, as errors can occur with the over-reliance on software (e.g., it should not be a substitute for critical thinking, Nuijten et al., 2016). Additionally, many tools assume data normality, place an overemphasis on statistical significance, and may rely on simplified assumptions that do not reflect the actual data. When computing sample size estimates, it is important to remember that these values are estimations, not exact calculations guaranteed to produce a specific result (Batterham \& Atkinson, 2005). For example, it is hard to accurately estimate all parameters from a study, and if any were incorrect, then the sample size estimate tied to that specific level of power may be incorrect (Albers \& Lakens, 2018).

Power analyses are not always the best tool for the job. Changes in publication practices and research design have also created new challenges in providing a sample size plan for a research study. While statistics courses often suggest that a specific research design leads to a specific statistical test, meta-science work has shown that given the same data and hypothesis, researchers can come up with multiple ways to analyze the data (Coretta et al., 2023; Silberzahn et al., 2018). Therefore, a single power analysis only corresponds to the specific analysis that the researcher expects to implement. Analyses may evolve during the research project or be subject to secondary analysis; thus, power and sample size estimation based on one analysis is potentially less useful than previously imagined. Further, research projects often have multiple testable hypotheses, but it is unclear which hypothesis or test should be used to estimate sample size with a power analysis. Last, research investigations may not even have a specific, testable hypothesis, as some projects are intended to curate a large dataset for future reuse (i.e., stimuli database creation, Buchanan et al., 2019).

In light of these analytical (or lack thereof) concerns, we propose a new method to determine a sample size in cases where a more traditional power analysis might be less appropriate or even impossible. This approach combines accuracy in parameter estimation (AIPE, Kelley, 2007; Maxwell et al., 2008) and bootstrapped simulation on pilot data (Rousselet et al., 2022). This method accounts for a potential lack of hypothesis test (or simply no good way to estimate an effect size of interest), and/or an exploratory design with an unknown set of potential hypotheses and analytical choices. Specifically this manuscript focuses on research designs that use multiple items to measure the phenomena of interest. For example, semantic priming is measured with multiple paired stimuli (Meyer \& Schvaneveldt, 1971), which traditionally has been analyzed by creating person or item-level averages to test using an ANOVA (Brysbaert \& Stevens, 2018). However, research implementing multilevel models with random effects for the stimuli used in a study has demonstrated potential variability in their impact on outcomes; thus we should be careful not to assume that all items in a research study have the same ``effect''.

@@

AIPE shifts the focus away from finding a significant p-value to finding a parameter that is ``accurately measured''. For example, researchers may wish to detect a specific mean in a study, \emph{M} = .35. They could then use AIPE to estimate the sample size needed to find a ``sufficiently narrow'' confidence interval around that mean. Sufficiently narrow is often defined by the researcher using a minimum parameter size of interest and confidence intervals. Therefore, they could decide that their 95\% confidence interval should be approximately between .20 and .50, and sufficiently narrow could be defined as a width of .30 or .15 on each side. While confidence intervals are related to null hypothesis significance testing (i.e., 95\% confidence intervals that do not include zero would indicate a significant difference from zero at \(\alpha\) \textless{} .05), AIPE procedures suggest how we can define a sample size with a given width of confidence interval, regardless of whether it includes zero.

@@

One form of data simulation is Bootstrapping, which involves using data obtained to simulate similar datasets by drawing from the original data with replacement (Efron, 2000; Rousselet et al., 2022). Bootstrapping allows one to calculate parameter estimates, confidence intervals, and to simulate the potential population distribution, shape, and bias. Simulation is often paired with re-creating a data set with a similar structure for testing analyses and hypotheses based on proposed effect sizes or suggested population means. Generally, we would suggest starting with pilot data of a smaller sample size (e.g., 20 to 50) to understand the variability in potential items used to represent your phenomenon, especially if they are to be used in a larger study. However, given some background knowledge about the potential items, one could simulate example pilot data to use in a similar manner in our suggested procedure. Pilot or simulated data would be used to estimate the variability within items and select a ``sufficiently narrow'' window for overall item confidence interval for AIPE (i.e., by selecting a specific standard error criterion, given the formula for confidence intervals). The advantage to this method over simple power estimation from pilot effect sizes is the multiple simulations to average out potential variability, as well as a shift away from traditional NHST to parameter estimation. Bootstrapping would then be used to determine how many participants may be necessary to achieve a dataset wherein as many items as required meet the pre-specified confidence interval.

@@

Researchers could then use sequential testing to estimate their parameter of interest after each participant's data or at regular intervals during data collection to determine whether they have achieved their expected width of the confidence interval around that parameter. One would set a minimum sample size (e.g.,, based on known data collection ability) and use the confidence interval width as a stopping rule (i.e., stop data collection when the CI is sufficiently narrow, as defined above). Next, researchers would use the estimated sample size associated with the simulation results of many items obtaining the stopping rule as a maximum sample size (e.g., , they expect 90\% of items to meet their stopping rule with 100 participants based on simulation). By defining each of these components, researchers could ensure a feasible minimum sample size, a way to stop data collection when goals have been met, and a maximum sample size rule to ensure an actual end to data collection. The maximum stopping rule could also be defined by resources (e.g., two semesters data collection), but nevertheless should be included. Therefore, we propose a method that leverages the ideas behind AIPE, paired with simulation and bootstrapping, to estimate the minimum and maximum proposed sample sizes and stopping rules for studies that use multiple items with expected variability in their estimates to measure an overall phenomena.

\hypertarget{proposed-method-for-sample-size-planning}{%
\subsection{Proposed Method for Sample Size Planning}\label{proposed-method-for-sample-size-planning}}

Building on these ideas, we suggest the following procedure to determine a sample size for each item:

\hypertarget{calculate-the-stopping-rule}{%
\subsubsection{Calculate the Stopping Rule}\label{calculate-the-stopping-rule}}

\begin{enumerate}
\def\labelenumi{\arabic{enumi})}
\item
  Use pilot data that closely resembles data you intend to collect. This dataset should contain items that are identical or similar to those that will be implemented in the study. In this procedure, it is important to ensure that the data is representative of a larger population of sampled items that you intend to assess. Generally, pilot data sample sizes will be smaller than the overall intended project (e.g., 20 to 50), as the goal would be to determine how many participants would be necessary to reach a ``stable'' standard error.
\item
  For each item in the pilot data, calculate the standard error (SE). Select a cutoff SE that defines when items are considered ``accurately measured''. The simulations described in the Data Simulation section will explore what criterion should be used to determine the cutoff SE from the pilot data.
\end{enumerate}

\hypertarget{bootstrap-samples}{%
\subsubsection{Bootstrap Samples}\label{bootstrap-samples}}

\begin{enumerate}
\def\labelenumi{\arabic{enumi})}
\setcounter{enumi}{2}
\tightlist
\item
  Sample, with replacement, from your pilot data using sample sizes starting at a value that you consider the minimal sample size per item and increase in small units up to a value that you consider the maximum sample size. We will demonstrate example maximum sample sizes based on the data simulation below; however, a practical maximum sample size may be determined by time (e.g., one semester data collection) or resources (e.g., 200 participants worth of funding). As for the minimal sample size, we suggest using 20 as a reasonable value for simulation purposes. For each sample size simulation, calculate the standard error for each item. Use multiple simulations (e.g., n = 500 to 1000) to avoid issues with random sampling variability.
\end{enumerate}

\hypertarget{determine-minimum-maximum-sample-size}{%
\subsubsection{Determine Minimum, Maximum Sample Size}\label{determine-minimum-maximum-sample-size}}

\begin{enumerate}
\def\labelenumi{\arabic{enumi})}
\setcounter{enumi}{3}
\item
  Use the simulated SEs to determine the percentage of items that meet the cutoff score determined in Step 2. Each sample size from Step 3 will have multiple bootstrapped simulations, and therefore, create an average percentage score for each sample size for Step 5.
\item
  Find the minimum sample size so that 80\%, 85\%, 90\%, and 95\% of the items meet the cutoff score and can be considered accurately measured. We recommend these scores to ensure that most items are accurately measured, in a similar vein to the common power-criterion suggestions. Each researcher can determine which of these is their minimum or maximum sample size (e.g., individuals can choose to use 80\% as a minimum and 90\% as a maximum or use values from Step 3 based on resources).
\end{enumerate}

\hypertarget{report-results}{%
\subsubsection{Report Results}\label{report-results}}

\begin{enumerate}
\def\labelenumi{\arabic{enumi})}
\setcounter{enumi}{5}
\tightlist
\item
  Report these values, and designate a minimum sample size, the cutoff/stopping rule criterion, and the maximum sample size. Each researcher should also report if they plan to use an adaptive design, which would stop data collection after meeting the cutoff criterion for each item.
\end{enumerate}

These steps are summarized in Table \ref{tab:table-summary}. We will first demonstrate the ideas behind the steps using open data (Balota et al., 2007; Brysbaert et al., 2014). This example will reveal a few areas of needed exploration for the steps. Next, we portray simulations for the proposed procedure and find solutions to streamline and improve the sample size estimation procedure. Table \ref{tab:table-summary} shows the results of the simulations and solutions on the right hand side. Finally, we include additional resources for researchers to use to implement the estimation procedure.

@@

In this section, we provide an example of the suggested procedure. The first dataset includes concreteness ratings from Brysbaert et al. (2014). Instructions given to participants denoted the difference between concrete (i.e., ``refers to something that exists in reality'') and abstract (i.e., ``something you cannot experience directly through your senses or actions'') terms. Participants were then asked to rate concreteness of terms using a 1 (\emph{abstract}) to 5 (\emph{concrete}) scale. This data represents a small scale dataset that could be used as pilot data for a study using concrete word ratings. The data is available at \url{https://osf.io/qpmf4/}.

The second dataset includes a large scale dataset with response latencies, the English Lexicon Project {[}ELP; Balota et al. (2007){]}. The ELP consists of lexical decision response latencies for written English words and pseudowords. In a lexical decision task, participants simply select ``word'' for real words (e.g., \emph{dog}) and ``nonword'' for pseudowords (e.g., \emph{wug}). The trial level data is available here: \url{https://elexicon.wustl.edu/}. Critically, in each of these datasets, the individual trial level data for each item is available to simulate and calculate standard errors on. Data that has been summarized could potentially be used, as long as the original standard deviations for each item were present. From the mean and standard deviation for each item, a simulated pilot dataset could be generated for estimating new sample sizes. All code to estimate sample sizes is provided on our OSF page, and this manuscript was created with a \emph{papaja} (Aust et al., 2022) formatted Rmarkdown document.

For this example, imagine a researcher who wants to determine the differences in response latencies for abstract and concrete words. They will select \emph{n} = 40 words from the rating data from Brysbaert et al. (2014) that are split evenly into abstract and concrete ends of the rating scale. In the experiment, each participant will be asked to rate the words for their concreteness, and then complete a lexical decision task with these words as the phenomenon of interest. Using both datasets and the procedure outlined above, we can determine the sample size necessary to ensure adequately measured concreteness ratings and response latencies.

\emph{Step 1}. The concreteness ratings data includes 27031 concepts that were rated for their concreteness. We randomly selected \emph{n} = 20 abstract words (\(M_{Rating}\) \textless= 2) and \emph{n} = 20 concrete words (\(M_{Rating}\) \textgreater= 4). In the original study, not every participant rated every word, which created uneven sample sizes for each word. Further, participants were allowed to indicate they did not know a word, and those responses were set to missing data. In our sample of 40 words, the average pilot sample size was 28.52 (\emph{SD} = 1.80), and we will use 29 as our pilot sample size for this example.

We first filtered the ELP data to the same real words as the concreteness subset selected above, and this data includes 27031 real words. The average pilot sample size for this random sample was 32.67 (\emph{SD} = 0.57), and \emph{n} = 33 will be our pilot size for this example.

@@

\emph{Step 3-5}. The pilot data was then simulated 500 times, with replacement, with samples of 20 to 300 participants per item increasing in units of 5, for concreteness ratings and lexical decision latencies separately (Step 3). The standard error of each item was calculated for the bootstrapped samples separately for concreteness ratings and lexical decision times (Step 4), and the percentage of items below each potential cutoff was gathered for each (Step 5). The smallest sample size with at least 80\%, 85\%, 90\%, and 95\% of items below the cutoff are reported in Table \ref{tab:table-example} for each task (Step 5).

@@

\caption{\label{tab:table-example}Sample Size Estimates by Decile for Example Study}
